---
title: "Data analysis"
author: "Albina Gibadullina"
date: "October 11, 2020"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

View available packages in CRAN: https://cran.r-project.org/web/views/


Information on current R version, operating system, loaded packages
```{r}
sessionInfo()
```


Install packages for network analysis
```{r}
#install.packages("igraph", "network", "sna", "tidyverse", "scales", "plotly")
```


Upload all neccessary packages
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(scales))
suppressPackageStartupMessages(library(plotly))
```


Update all packages
```{r}
#update.packages()
```


To unload a package
```{r}
#detach("package:ggplot2", unload=TRUE)
```


To remove a package
```{r}
#remove.packages("ggplot2")
```


Get help on a package
```{r}
help(package = "ggplot2")

#Get additional info for packages
browseVignettes("ggplot2")
```


Update R version
```{r}
# install.packages("installr")
# library(installr)
# updateR()
```


Styles:
**bold**
*italics*
# Header1
## Header2
- bullet point 1


## R Markdown

Binomial distribution
```{r}
#Assign values to variables
n <- 500
p <- 0.1
X <- 50

#computes FY(y)
pbinom(X,n,p)

#computes fY(y)
dbinom(X,n,p) 
```


Normal approximation to the Binomial
```{r}
# generate m i.i.d. binomial random variables
## rbinom - generation of a binomial distribution with specified paramters
p = 0.5; n = 100; m=10000;
yn = rbinom(m,n,p);

# standardize these variables
zn = (yn - n*p)/sqrt(p*(1-p)*n);

# draw the histogram
hist(zn,100,main=paste("Histogram of Binomial Distribution"))
```


Exponential distribution
```{r}
#Assign values to variables
y <- 5
lambda <- 2

#computes FY(y)
pexp(y, lambda)

#computes fY(y)
dexp(y, lambda) 
```


Poisson distribution
```{r}
#Assign values to variables
y <- 5
lambda <- 2

#computes FY(y)
ppois(y, lambda)

#computes fY(y)
dpois(y, lambda) 
```


Normal distribution
```{r}
#creates a vector (-3.00,-2.99,...,2.99,3.00)
z = seq(-3,3,0.01)

#computes the normal density at each point of z
fz = dnorm(z)

#plots z against fz (as a line)
plot(z, fz, type="l")

#Assign values to variables
X <- 0.5
Mean <- 0 
SD <- 1

#computes FY(y)
pnorm(X, Mean, SD, lower.tail=FALSE)

#computes fY(y)
dnorm(X, Mean, SD)

# set seed to reproduce the results
set.seed(123); 

# generates n=3 random deviates, normal dist, finds mean

y1=rnorm(3,mean=2,sd=1); mean(y1);
y2=rnorm(3,mean=2,sd=1); mean(y2);
y3=rnorm(3,mean=2,sd=1); mean(y3);
y4=rnorm(3,mean=2,sd=1); mean(y4);
y5=rnorm(3,mean=2,sd=1); mean(y5);
```


Chi-square distribution
```{r}

# df is the number of degrees of freedom
z = seq(0.1,8,0.01);
fz1 = dchisq(z,df=1);
fz2 = dchisq(z,df=2);
fz3 = dchisq(z,df=3);

# plot distribution functions for different df
plot(z, fz1, type="l");
lines(z, fz2, col="green");
lines(z, fz3, col="red");

#Assign values to variables
CI<- 0.95
df <- 10

#computes FY(y)
qchisq(CI, df)

#computes fY(y)
dchisq(CI,df)
```


T-distribution
```{r}
z = seq(-3,3,0.01);
fz1 = dnorm(z);
fz2 = dt(z,df=2);
plot(z, fz1, type="l");
lines(z, fz2, type="l",col="green")
```

