---
title: "Data analysis"
author: "Albina Gibadullina"
date: "October 11, 2020"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set Up

[View available packages in CRAN](https://cran.r-project.org/web/views/) 


Information on current R version, operating system, loaded packages
```{r}
#sessionInfo()
```


Install packages for network analysis
```{r}
#install.packages("igraph", "network", "sna", "tidyverse", "scales", "plotly")
```


Upload all neccessary packages
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(scales))
suppressPackageStartupMessages(library(plotly))
```


Update all packages
```{r}
#update.packages()
```


To unload a package
```{r}
#detach("package:ggplot2", unload=TRUE)
```


To remove a package
```{r}
#remove.packages("ggplot2")
```


Get help on a package
```{r}
# help(package = "ggplot2")

#Get additional info for packages
# browseVignettes("ggplot2")
```


Update R version
```{r}
# install.packages("installr")
# library(installr)
# updateR()
```



## R Markdown Styles

**bold**
*italics*
~~strikethrough~~


# Header1
## Header2
### Header 3



#### Unordered list

* Item 1
* Item 2
    + Item 2a
    + Item 2b

#### Ordered list
1. Item 1
2. Item 2
3. Item 3
    + Item 3a
    + Item 3b


[linked phrase](http://example.com)

Inline equation:
$5x+2$

Display equation:
$$ 5x+2 $$

## Inputting data

```{r}
x <- 1
print(x) #explict printing
msg <- "hello"
msg #auto-printing

#Create a sequence of values
y <- 1:20
```


Creating vectors

```{r}
#Numeric data
x <- 1 # x equals 1
x <- 1L # specify that 1 is an integer

# infinity - Inf
1/Inf
```


Creating Vectors
```{r}
# The c() function can be used to create vectors of objects

x1 <- c(0.5, 0.6) ## numeric
x2 <- c(TRUE, FALSE) ## logical
x3 <- c(T, F) ## logical
x4 <- c("a", "b", "c") ## character
x5 <- 9:29 ## integer
x6 <- c(1+0i, 2+4i) ## complex

# Using the vector() function
x7 <- vector("numeric", length = 10)
x1; x2; x3; x4; x5; x6; x7
```


Vectorized Operations
```{r}
#Many operations in R are vectorized making code more efficient, concise, and easier to read.
x <- 1:4 
y <- 6:9
x + y

#Make logical vectors
x > 2
x >= 2
y == 8

x*y
x/y

#Vectorized Matrix Operations
x <- matrix(1:4, 2, 2)
y <- matrix(rep(10, 4), 2, 2)

x*y ## element-wise multiplication
x/y  ## element-wise division
x %*% y ## true matrix multiplication
```


Mixing Objects
```{r}
#When different objects are mixed in a vector, coercion occurs so that every element in the vector is of the same class.

y <- c(1.7, "a") ## convert 1.7 to character
y <- c(TRUE, 2) ## convert TRUE to numeric = 1
y <- c("a", TRUE) ## convert TRUE to character
```


Explicit Coercion
```{r}
#Objects can be explicitly coerced from one class to another using the as.* functions, if available.

x <- 0:6
class(x)
as.numeric(x)
as.logical(x)
as.character(x)

#Nonsensical coercion results in NAs:

x <- c("a", "b", "c")
as.numeric(x)
as.logical(x)
as.complex(x)
```


Matrices
```{r}
#Matrices are vectors with a dimension attribute. The dimension attribute is itself an integer vector of length 2 (nrow, ncol)

m <- matrix(nrow = 2, ncol = 3)
m
dim(m)
attributes(m)

#Matrices are constructed column-wise, so entries can be thought of starting in the “upper left” corner and running down the columns.
m <- matrix(1:6, nrow = 2, ncol = 3)
m

#Matrices can also be created directly from vectors by adding a dimension attribute.
m <- 1:10
m

dim(m) <- c(2, 5)
m

# cbind-ing and rbind-ing
## Matrices can be created by column-binding or row-binding with cbind() and rbind()

x <- 1:3
y <- 10:12
cbind(x, y)

rbind(x, y)
```


Lists
```{r}
#Lists are a special type of vector that can contain elements of different classes

x <- list(1, "a", TRUE, 1 + 4i)
x
```


Factors
```{r}
#Factors are used to represent categorical data. Factors can be unordered or ordered

x <- factor(c("yes", "yes", "no", "yes", "no"))
x

# frequency count summary
table(x)

# turn categorical data into integer data (how R sees categorical data)
unclass(x)

#The order of the levels can be set using the levels argument to factor(). R lists levels alphabetically unless specified

x <- factor(c("yes", "yes", "no", "yes", "no"),
 levels = c("yes", "no"))
x
```


Missing values
```{r}
# NA denotes missing values
# is.na() is used to test objects if they are NA

# NaN devotes undefined mathematical operations
# is.nan() is used to test for NaN

x <- c(1, 2, NA, 10, 3)
is.na(x)
is.nan(x)

# A NaN value is also NA but the converse is not true

x <- c(1, 2, NaN, NA, 4)
is.na(x)
is.nan(x)
```


Removing NA Values
```{r}
#A common task is to remove missing values (NAs).
x <- c(1, 2, NA, 4, NA, 5)

#extract vector with missing values
bad <- is.na(x)

#exclude vector with all the missing values
x[!bad]


#What if there are multiple things and you want to take the subset with no missing values?
x <- c(1, 2, NA, 4, NA, 5)
y <- c("a", NA, "b", "d", NA, "f")

#Leave values that are not missing for both x and y
good <- complete.cases(x, y)
good
x[good]
y[good]
```



Data Frames
```{r}
# Data frames are used to store tabular data. Unlike matrices, data frames can store different classes of objects in each column (just like lists). 

x <- data.frame(id = c("A", "B", "C", "D"), foo = 1:4, bar = c(T, T, F, F))

# row.names lists all the rows
row.names(x)

# measure number of rows
nrow(x)

# measure number of columns
ncol(x)

# read.table() or read.csv() to create a data frame

# data.matrix() to converte data frame to a matrix, coerces objects so that they are all the same!
```


Names
```{r}
#R objects can also have names, which is very useful for writing readable code and self-describing objects.

x <- 1:3
names(x) <- c("foo", "bar", "norf")

x
names(x)

#Lists can also have names.
x <- list(a = 1, b = 2, c = 3)
x
names(x)

#Matrics can also have names.
m <- matrix(1:4, nrow = 2, ncol = 2)
dimnames(m) <- list(c("a", "b"), c("c", "d"))
m
```


## Reading data

There are a few principal functions reading data into R

* read.table, read.csv
    + for reading tabular data, read.csv is identical to read.table except that the default separator is a comma.
* readLines
    + for reading lines of a text file
* source
    + for reading in R code files (inverse of dump)
* dget
    + for reading in R code files (inverse of dput)
* load
    + for reading in saved workspaces
* unserialize
    + for reading single R objects in binary form


## Writing Data

There are analogous functions for writing data to files

* write.table
    + for writing tabular data
* writeLines
    + for writing lines of a text file
* dump
    + for writing in R code files
* dput
    + for writing in R code files (inverse of dput)
* save
    + for saving in workspaces
* serialize
    + for writing single R objects in binary form


### read.table arguments:
* file, the name of a file, or a connection
* header, logical indicating if the file has a header line
* sep, a string indicating how the columns are separated
* colClasses, a character vector indicating the class of each column in the dataset
* nrows, the number of rows in the dataset
* comment.char, a character string indicating the comment character
* skip, the number of lines to skip from the beginning
stringsAsFactors, should character variables be coded as factors?


#### Reading in Large Datasets with read.table

Use the colClasses argument - specifying this option instead of using the default can make ’read.table’ run MUCH faster, often twice as fast.

If all of the columns are “numeric”:
```{r}
initial <- read.table("datatable.txt", 
                      colClasses = "numeric")
```


If you don't know the classes of each column:
```{r}
# Read the first nrows and identify the class of data for each column
initial <- read.table("datatable.txt", 
                      nrows = 100)
classes <- sapply(initial, class)

# Read the dataset telling R the data class for each column
tabAll <- read.table("datatable.txt",
                     colClasses = classes)
```



Extracting data from a dataframe
```{r}
# Create a dataset
dataset1 <- data.frame(c(20,40,60),c(10,5,7),c(1,2,3))
names(dataset1) <- c("Height", "Weight", "Age")

# Extract all rows, all colums
dataset1[,]

# Extract first row and all columns
dataset1[1,]

# Extract first two rows and all columns
dataset1[1:2,]

# Extract first and third row and all columns
dataset1[c(1,3),]

# Extract first row and second and third column
dataset1[1,2:3]

# Extract third column with all rows
dataset1[,3]

# Extract first and third column with all rows
dataset1[,c(1,3)]

# Extract element at 2nd row, 3rd column
dataset1[2,3]
```



Subsetting Lists
```{r}
#Create a list
x <- list(foo = 1:4, bar = 0.6)

# Extract 1st element sane class as the original - list
x[1]

# Extract the value of the 1st element
x[[1]]

# Extract the value of the the element "bar"
x$bar

# Same as before
x[["bar"]]

# Extract 2nd element "bar"
x["bar"]


# Extract more than one element
x <- list(foo = 1:4, bar = 0.6, baz = "hello")
x[c(1, 3)]


#Subsetting Nested Elements of a List
x <- list(a = list(10, 12, 14), b = c(3.14, 2.81))
x[[c(1, 3)]]
x[[c(2, 1)]]

```


Subsetting a Matrix
```{r}
x <- matrix(1:6, 2, 3)
x[1, 2]
x[2, 1]
x[1, ]
x[, 2]

#By default, when a single element of a matrix is retrieved, it is returned as a vector of length 1 rather than a 1 × 1 matrix. This behavior can be turned off by setting drop = FALSE.
x <- matrix(1:6, 2, 3)
x[1, 2]
x[1, 2, drop = FALSE]

# Similarly, subsetting a single column or a single row will give you a vector, not a matrix (by default).
x <- matrix(1:6, 2, 3)
x[1, ]
x[1, , drop = FALSE]
```


HW1, Week1
```{r}
library(readr)

# Read the csv file
hw1_data <- read.csv("hw1_data.csv")

# List the column names
colnames(hw1_data)

# Extract 2 top rows
hw1_data_2toprows <- head(hw1_data, 2)
print(hw1_data_2toprows)

# Count the number of rows
nrow(hw1_data)

# Extract 2 bottom rows
hw1_data_2bottomrows <- tail(hw1_data, 2)
print(hw1_data_2bottomrows)

# Extract row#47, column#1
hw1_data[47,1]

# Count missing values in column 1
table(is.na(hw1_data[,1])) 

#create a subset of data feauturing column 1 excluding all missing values
sub <- subset(hw1_data, !is.na(hw1_data[,1]), select =Ozone) 

# Find average of column 1 excluding missing values
apply(sub,2,mean)


# Turn the data file into df
hw1_data_df <- data.frame(hw1_data)

# Extract the subset of rows of the data frame where Ozone values are above 31 and Temp values are above 90. What is the mean of Solar.R in this subset?

sub1 = subset(hw1_data, Ozone > 31 & Temp > 90, select = Solar.R)
apply(sub1, 2, mean)


# What is the mean of "Temp" when "Month" is equal to 6?
sub2 = subset(hw1_data, Month==6, select = Temp)
apply(sub2, 2, mean)

# What was the maximum ozone value in the month of May (i.e. Month is equal to 5)?

sub3 = subset(hw1_data, Month==5 & !is.na(Ozone), select = Ozone)
apply(sub3, 2, max)
  
```


### Textual formats
dumping and dputing are useful because the resulting textual format is edit-able, and in the case of corruption, potentially recoverable.

dput-ting R Objects
```{r}
#Another way to pass data around is by deparsing the R object with dput and reading it back in using dget.
y <- data.frame(a = 1, b = "a")
dput(y)

#Create a new file with this information
dput(y, file = "y.R")

# Read it into R
new.y <- dget("y.R")

# Create an output that is readable
new.y
```


Dumping R Objects
```{r}
#Multiple objects can be deparsed using the dump function and read back in using source.
x <- "foo"
y <- data.frame(a = 1, b = "a")
dump(c("x", "y"), file = "data.R")
rm(x, y)
source("data.R")
x
y
```


Reading Lines of a Text File
```{r}
con <- url("http://www.jhsph.edu", "r")
x <- readLines(con)
head(x)
```




## Distributions


Binomial distribution
```{r}
#Assign values to variables
n <- 500
p <- 0.1
X <- 50

#computes FY(y)
pbinom(X,n,p)

#computes fY(y)
dbinom(X,n,p) 
```


Normal approximation to the Binomial
```{r}
# generate m i.i.d. binomial random variables
## rbinom - generation of a binomial distribution with specified paramters
p = 0.5; n = 100; m=10000;
yn = rbinom(m,n,p);

# standardize these variables
zn = (yn - n*p)/sqrt(p*(1-p)*n);

# draw the histogram
hist(zn,100,main=paste("Histogram of Binomial Distribution"))
```


Exponential distribution
```{r}
#Assign values to variables
y <- 5
lambda <- 2

#computes FY(y)
pexp(y, lambda)

#computes fY(y)
dexp(y, lambda) 
```


Poisson distribution
```{r}
#Assign values to variables
y <- 5
lambda <- 2

#computes FY(y)
ppois(y, lambda)

#computes fY(y)
dpois(y, lambda) 
```


Normal distribution
```{r}
#creates a vector (-3.00,-2.99,...,2.99,3.00)
z = seq(-3,3,0.01)

#computes the normal density at each point of z
fz = dnorm(z)

#plots z against fz (as a line)
plot(z, fz, type="l")

#Assign values to variables
X <- 0.5
Mean <- 0 
SD <- 1

#computes FY(y)
pnorm(X, Mean, SD, lower.tail=FALSE)

#computes fY(y)
dnorm(X, Mean, SD)

# set seed to reproduce the results
set.seed(123); 

# generates n=3 random deviates, normal dist, finds mean

y1=rnorm(3,mean=2,sd=1); mean(y1);
y2=rnorm(3,mean=2,sd=1); mean(y2);
y3=rnorm(3,mean=2,sd=1); mean(y3);
y4=rnorm(3,mean=2,sd=1); mean(y4);
y5=rnorm(3,mean=2,sd=1); mean(y5);
```


Chi-square distribution
```{r}

# df is the number of degrees of freedom
z = seq(0.1,8,0.01);
fz1 = dchisq(z,df=1);
fz2 = dchisq(z,df=2);
fz3 = dchisq(z,df=3);

# plot distribution functions for different df
plot(z, fz1, type="l");
lines(z, fz2, col="green");
lines(z, fz3, col="red");

#Assign values to variables
CI<- 0.95
df <- 10

#computes FY(y)
qchisq(CI, df)

#computes fY(y)
dchisq(CI,df)
```


T-distribution
```{r}
z = seq(-3,3,0.01);
fz1 = dnorm(z);
fz2 = dt(z,df=2);
plot(z, fz1, type="l");
lines(z, fz2, type="l",col="green")
```

